Detecting confounder and center bias in machine learning models

### 2021-10-28 manuscript.tex
fix
Updates from Overleaf

Predictive modelling has recently become increasingly important in biomedical research and holds promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized, spisak2020pain, walsh2021dome}. When evaluating the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) is far from being the only important consideration. Biomedical validity (i.e. whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are crucial requirements for candidate biomarkers \citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.

Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases \citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability. Examples include models picking up on measurement artifacts (in-scanner head motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's \citep{rao2017predictive}, attention deficit hyperactivity disorder \citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD) \citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence \citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance \citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (racially biased machine learning models \citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.

While various data cleaning methods might help in mitigating confounder bias \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive, johnson2007adjusting}, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.

Powerful and robust statistical tests to quantify confounding effects in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches. Although confounder bias can be expressed in terms of conditional independence of the model predictions on the target and the confoudner variables, the proper evaluation of the conditional independence among these variables is challenging. Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation) are not valid measures of conditional independence. Although warnings about this issue were given from early on \citep{korn1984ranges}, and received a fair amount of attention recently \citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g. feature-set characteristics and model regularization \citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.

Alternative techniques for quantifying confoudner bias either fail to control type error (as  the known in the case of balanced permutations \cite{southworth2009properties, hemerik2018exact}, used in \cite{chaibub2019permutation}), or do not aim at all to do so \cite{ferrari2020measuring, wachinger2021detect}. 
Moreover, without some modifications, they are applicable for either only numeric \cite{wachinger2021detect} or only categorical variables \citep{chaibub2019permutation, ferrari2020measuring} and involve re-fitting the model \citep{chaibub2019permutation, ferrari2020measuring}, which might not be feasible for models with high a computational cost (e.g. with nested cross-validation). Some techniques \cite{wachinger2021detect}, furthermore, probe a null hypothesis (direct causation between the features and the prediction target) that is not applicable for for many kinds of data and models \footnote{For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation. In such cases, the alternative hypothesis of this test might be true in absence of any real confounder. }.

@@

%In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}. This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem). 
%As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g. \citep{korn1984ranges} or Fig. \ref{fig:sim-h0-demo} in this paper), as commonly seen in machine learning.

%To overcome these limitations, I extend the permutation-based conditional independence framework of \cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output. 
%The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
%Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
%The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).

%The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models. 
%First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
%Second, functional connectivity data from the ABIDE \citep{di2014autism} database is used to investigate the potential motion- and center-bias \cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.

@@

The idea of conditional independence provides a straightforward framework for assessing the confounder bias in predictive models. However, handling the non-normal and/or non-linear conditional dependence often seen in predictive models \citep{garcia2009study, kristensen2017whole} poses a great challenge.
In fact, as recently shown by Shah and colleagues \cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power. In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios. For instance, a trivial candidate, partial Spearman correlation, exhibits inflated type I errors even with slight violations of normality and linearity, as clearly demonstrated with simulated data on Fig. \ref{fig:sim-h0-demo}.

As the most problematic from the three variables (target, prediction, confounder) is clearly the model prediction \citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics. Exactly this can be achieved with the combination of the novel framework of conditional permutation testing (CPT) \cite{berrett2020conditional} and the use of generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}-based conditional distribution estimation. The proposed approach offers two novel tests for probing confounder bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
These tests place no assumptions on the conditional distributions of the model predictions, ensuring valid model diagnostics even in cases of non-normally and non-linearly dependent predictions.

@@

Functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.

@@

Functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g. \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

@@

The assessment of confoudner bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings. However, statistical testing of conditional independence is challenging: analogs of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity \citep{korn1984ranges} (see an example on Fig. \ref{fig:sim-h0-demo}). While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e. the nonexistence of a uniformly valid conditional independence test) \cite{shah2020hardness} clearly implies that conditional independence-based confoudner bias testing must be designed so that its suitability for the particular problem may be judged easily.

In this paper, I have proposed a solution for confounder testing, which was designed to be suitable for the specific characteristics of the 'confoudner problem' in predictive modelling. Namely, the proposed approach places assumptions only about the conditional distribution of the confoudner and the target variables, as required by the applied generalized additive \cite{hastie1987generalized} or multinomial logistic model \cite{bennett1966multiple, jones1975proability}. In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analysis, as well. Accordingly, usually there is sufficient expert knowledge about the applicability of the required assumptions and data acquisition is probably already designed so that the assumptions are likely fulfilled.
Of key importance, the proposed solution is \emph{absolutely assumption-free} regarding the distribution and dependency of the prediction output, which is, in most of the cases, unknown and can often be expected to be non-normal and/or non-linear (due to e.g model regularization) \citep{garcia2009study, kristensen2017whole}. 
This property distinguishes the approach from other alternatives as it guarantees a valid type-I error control even in cases of non-normally and non-linearly dependent predictions, i.e. in cases where Pearson and Spearman partial correlations, and many other methods fail.

@@

%In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g. \citep{spisak2020pain}. However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable. In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association. 
%Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).
%These examples clearly indicate that guiding model diagnostics based on unconditioned associations between the confounder and the model output might result in overly strict decisions. More realistic is to aim at models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e. the prediction-confounder association is proportional to, and can be fully explained by, the target-confounder association.
%Put simply, this is what is tested by the proposed partial confounder test.

@@

A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making \citep{woo2017building, wager2013fmri, spisak2020pain}. However, the presence of confounders typical for biomedical research (e.g. sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g. imaging artifacts) presents a great challenge to these efforts.
The usefulness of the proposed tests is demonstrated in two such examples, using the HCP \citep{van2013wu}  and the ABIDE \citep{di2014autism} datasets.
%First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
$Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g. \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
The hypothesised confounder effects were clearly present in both datasets.

In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings \citep{dubois2018distributed, lohmann2021predicting}) could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed bias. Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time). To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g. slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff). This is obviously undesirable for any brain-based predictive model of fluid intelligence.

In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD. For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g. the CMU (Carnegie Mellon University) center as neurotypical control participants. Similarly, the models biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion, e.g. due to atypical motor function (as known e.g. for ADHD \citep{eloyan2012automated}).

@@

Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign. Meanwhile, the prediction performance substantially dropped (although still remained statistically significant). 
A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
This effect might be boosted by the previously reported "residual motion artifacts" and their regional interactions \citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
 As COMBAT was originally developed for harmonizing effects of categorical variables (e.g. center or batch), its application for continuous confounder variables is not trivial. The presented method of inputting the binned "motion groups" into COMBAT resulted in an efficient removal of motion-bias and, at the same time, preserved a significant amount of signal-of-interest. 
 Nevertheless, inputting discretized versions of continuous variables into COMBAT might be sub-optimal and raises further questions e.g regarding the optimal number of bins used during the discretization.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%TC:ignore
\section{Methods}
\label{sec:methods}

@@

 Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, fiedler2011mediation, peters2016causal, candes2016panning}. Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
 Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional independence testing is effectively impossible. In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provide a valid type I error control and, at the same time, a non-trivial statistical power.

@@

Recently, \cite{candes2016panning} and \cite{berrett2020conditional} have demonstrated that valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
Specifically, the conditional permutation test (CPT) of Berrett and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable. Thereby, it incorporates the information available on the conditional distribution of interest into the permutation-based inference in a statistically valid manner. Moreover, as shown in this paper, CPT can be combined with a generalized additive model (GAM) to handle non-linear relationships, rendering it as an appealing candidate for testing confounder effects in predictive modelling.

As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder. In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether the features $X$ still affect $Y$, when controlling for $Z$.
For instance, in genome-wide association studies, CPT can be used to determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.

@@

As shown in \citep{berrett2020conditional}, to ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:

@@

As long as the numerator of Eq. \ref{eq-pperm} is non-zero for all $c_\pi \in C$ and $y \in Y$, the conditional permutations constitute an algebraic group, thus, as shown by \citep{hemerik2018exact}, an unbiased estimate of the p-value under the null can be obtained as:
$$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$

While the group property of the conditioned permutations provides a straightforward proof for the validity of the approach, for an alternative verification see the proof of Theorem 1 in \citep{berrett2020conditional}.

>>>>

Predictive modelling has recently become increasingly important in biomedical research and holds promise for delivering biomarkers that substantially impact clinical practice and public health\citep{kent2018personalized, spisak2020pain, walsh2021dome}. When evaluating the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) is far from being the only important consideration. Biomedical validity (i.e. whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are crucial requirements for candidate biomarkers\citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.

Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases\citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability. Examples include models picking up on measurement artifacts (in-scanner head motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's\citep{rao2017predictive}, attention deficit hyperactivity disorder\citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD)\citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence\citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance\citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (racially biased machine learning models\citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.

While various data cleaning methods might help in mitigating confounder bias\citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive, johnson2007adjusting}, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest\citep{wachinger2021detect}.

Powerful and robust statistical tests to quantify confounding effects in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches. Although confounder bias can be expressed in terms of conditional independence of the model predictions on the target and the confoudner variables, the proper evaluation of the conditional independence among these variables is challenging. Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation) are not valid measures of conditional independence. Although warnings about this issue were given from early on\citep{korn1984ranges}, and received a fair amount of attention recently\citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g. feature-set characteristics and model regularization\citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.

Alternative techniques for quantifying confoudner bias either fail to control type error (as  the known in the case of balanced permutations\cite{southworth2009properties, hemerik2018exact}, used in\cite{chaibub2019permutation}), or do not aim at all to do so\cite{ferrari2020measuring, wachinger2021detect}. 
Moreover, without some modifications, they are applicable for either only numeric\cite{wachinger2021detect} or only categorical variables\citep{chaibub2019permutation, ferrari2020measuring} and involve re-fitting the model\citep{chaibub2019permutation, ferrari2020measuring}, which might not be feasible for models with high a computational cost (e.g. with nested cross-validation). Some techniques\cite{wachinger2021detect}, furthermore, probe a null hypothesis (direct causation between the features and the prediction target) that is not applicable for for many kinds of data and models \footnote{For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation. In such cases, the alternative hypothesis of this test might be true in absence of any real confounder. }.

@@

%In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing\citep{dawid1979conditional}. This allows relating the problem to the seminal work of\cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem). 
%As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g.\citep{korn1984ranges} or Fig. \ref{fig:sim-h0-demo} in this paper), as commonly seen in machine learning.

%To overcome these limitations, I extend the permutation-based conditional independence framework of\cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output. 
%The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
%Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
%The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).

%The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models. 
%First, functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
%Second, functional connectivity data from the ABIDE\citep{di2014autism} database is used to investigate the potential motion- and center-bias\cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.

@@

The idea of conditional independence provides a straightforward framework for assessing the confounder bias in predictive models. However, handling the non-normal and/or non-linear conditional dependence often seen in predictive models\citep{garcia2009study, kristensen2017whole} poses a great challenge.
In fact, as recently shown by Shah and colleagues\cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power. In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios. For instance, a trivial candidate, partial Spearman correlation, exhibits inflated type I errors even with slight violations of normality and linearity, as clearly demonstrated with simulated data on Fig. \ref{fig:sim-h0-demo}.

As the most problematic from the three variables (target, prediction, confounder) is clearly the model prediction\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics. Exactly this can be achieved with the combination of the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} and the use of generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}-based conditional distribution estimation. The proposed approach offers two novel tests for probing confounder bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
These tests place no assumptions on the conditional distributions of the model predictions, ensuring valid model diagnostics even in cases of non-normally and non-linearly dependent predictions.

@@

Functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.

@@

Functional connectivity data from the ABIDE\citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

@@

The assessment of confoudner bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings. However, statistical testing of conditional independence is challenging: analogs of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity\citep{korn1984ranges} (see an example on Fig. \ref{fig:sim-h0-demo}). While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e. the nonexistence of a uniformly valid conditional independence test)\cite{shah2020hardness} clearly implies that conditional independence-based confoudner bias testing must be designed so that its suitability for the particular problem may be judged easily.

In this paper, I have proposed a solution for confounder testing, which was designed to be suitable for the specific characteristics of the 'confoudner problem' in predictive modelling. Namely, the proposed approach places assumptions only about the conditional distribution of the confoudner and the target variables, as required by the applied generalized additive\cite{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}. In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analysis, as well. Accordingly, usually there is sufficient expert knowledge about the applicability of the required assumptions and data acquisition is probably already designed so that the assumptions are likely fulfilled.
Of key importance, the proposed solution is \emph{absolutely assumption-free} regarding the distribution and dependency of the prediction output, which is, in most of the cases, unknown and can often be expected to be non-normal and/or non-linear (due to e.g model regularization)\citep{garcia2009study, kristensen2017whole}. 
This property distinguishes the approach from other alternatives as it guarantees a valid type-I error control even in cases of non-normally and non-linearly dependent predictions, i.e. in cases where Pearson and Spearman partial correlations, and many other methods fail.

@@

%In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.\citep{spisak2020pain}. However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable. In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association. 
%Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).
%These examples clearly indicate that guiding model diagnostics based on unconditioned associations between the confounder and the model output might result in overly strict decisions. More realistic is to aim at models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e. the prediction-confounder association is proportional to, and can be fully explained by, the target-confounder association.
%Put simply, this is what is tested by the proposed partial confounder test.

@@

A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making\citep{woo2017building, wager2013fmri, spisak2020pain}. However, the presence of confounders typical for biomedical research (e.g. sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g. imaging artifacts) presents a great challenge to these efforts.
The usefulness of the proposed tests is demonstrated in two such examples, using the HCP\citep{van2013wu}  and the ABIDE\citep{di2014autism} datasets.
%First, functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
%Second, functional connectivity data from the ABIDE\citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
%The hypothesised confounder effects were clearly present in both datasets.

In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings\citep{dubois2018distributed, lohmann2021predicting}) could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed bias. Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time). To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g. slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff). This is obviously undesirable for any brain-based predictive model of fluid intelligence.

In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD. For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g. the CMU (Carnegie Mellon University) center as neurotypical control participants. Similarly, the models biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion, e.g. due to atypical motor function (as known e.g. for ADHD\citep{eloyan2012automated}).

@@

Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign. Meanwhile, the prediction performance substantially dropped (although still remained statistically significant). 
A possible explanation to these phenomena is that in-scanner motion, as previously described\citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
This effect might be boosted by the previously reported "residual motion artifacts" and their regional interactions\citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
 As COMBAT was originally developed for harmonizing effects of categorical variables (e.g. center or batch), its application for continuous confounder variables is not trivial. The presented method of inputting the binned "motion groups" into COMBAT resulted in an efficient removal of motion-bias and, at the same time, preserved a significant amount of signal-of-interest. 
 Nevertheless, inputting discretized versions of continuous variables into COMBAT might be sub-optimal and raises further questions e.g regarding the optimal number of bins used during the discretization.

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%TC:ignore
\section{Methods}
\label{sec:methods}

@@

 Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications\citep{spirtes2000causation, fiedler2011mediation, peters2016causal, candes2016panning}. Recently,\cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
 Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional independence testing is effectively impossible. In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provide a valid type I error control and, at the same time, a non-trivial statistical power.

@@

Recently,\cite{candes2016panning} and\cite{berrett2020conditional} have demonstrated that valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
Specifically, the conditional permutation test (CPT) of Berrett and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable. Thereby, it incorporates the information available on the conditional distribution of interest into the permutation-based inference in a statistically valid manner. Moreover, as shown in this paper, CPT can be combined with a generalized additive model (GAM) to handle non-linear relationships, rendering it as an appealing candidate for testing confounder effects in predictive modelling.

As many related papers, the work of\cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder. In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether the features $X$ still affect $Y$, when controlling for $Z$.
For instance, in genome-wide association studies, CPT can be used to determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.

@@

As shown in\citep{berrett2020conditional}, to ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:

@@

As long as the numerator of Eq. \ref{eq-pperm} is non-zero for all $c_\pi \in C$ and $y \in Y$, the conditional permutations constitute an algebraic group, thus, as shown by\citep{hemerik2018exact}, an unbiased estimate of the p-value under the null can be obtained as:
$$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$

While the group property of the conditioned permutations provides a straightforward proof for the validity of the approach, for an alternative verification see the proof of Theorem 1 in\citep{berrett2020conditional}.
